Something about Performance, 2D Component of NEMO. (TBC)

### 2015-02-12 docs/papers/nemolite2d_performance/nemolite2d_perf.tex
WIP on describing the in-lining of Mom. kernels

@@

Before applying any code transformations, we benchmark the original
version of the code. We also benchmark the vanilla, unoptimised
version after it has been re-structured following the \psykal
approach.

{\bf Performance analysis as motivation for the code
  tranformations. More detail on each transformation.}

\begin{table}[!t]
\caption{The performance profile of the original and \psykal 
versions of nemolite2d ($128^{2}$ domain and Intel compiler).}
\label{TABLE_profile}
\centering
\begin{tabular}{l|c|c|c|c}
\hline
Section & \multicolumn{2}{c|}{Original} & \multicolumn{2}{c}{\psykal} \\
        & Time (s) & \%  & Time (s) & \% \\
\hline
Continuity  & 0.096 & 2.7  & 0.33  & 6.6 \\
Momentum    & 3.00  & 84.5 & 4.05  & 79.5 \\
BCs         & 0.16  & 4.6  & 0.29  & 5.7 \\
Time-update & 0.29  & 8.1  & 0.41  & 8.1 \\
\hline
\end{tabular}
\end{table}

Beginning with the vanilla \psykal version, we then apply a series of
code transformations while obeying the \psykal separation of concerns,
{\it i.e.} optimisation is restricted to the middle, {PS}y layer and leaves
the kernel and algorithm layers unchanged. The aim of these
optimisations is to recover, as much as is possible, the structure and
thus performance of the original version of the code. The
transformations we have performed are as follows:
\begin{enumerate}

\item Addition of 'safe-loop' directives

\item In-line Momentum kernel bodies into middle-layer code

\item Force SIMD vectorisation of the Momentum kernels using directives

\item Work around limitations related to derived types

\item In-line the Continuity kernel

\item In-line remaining kernels (BCs and time-update)

\item In-line field-copy operations

\end{enumerate}

>>>>

{\bf Description of the structure of the program at the Algorithm
  level so that it links in to Profile data in results section.}

@@

Before applying any code transformations, we benchmark the original
version of the code. We also benchmark the vanilla, unoptimised
version after it has been re-structured following the \psykal
approach. In addition to this benchmarking, we profile these versions
of the code at the algorithm level (using a high-level timing
API). The resulting profiles are given in
Table~\ref{TABLE_profile}. The Momentum section dominates the profile
of both versions of the code, accounting for around 80\% of the
wall-clock time spent doing time-stepping. In terms of performance
loss when moving from the original to the \psykal version of the code
however, the Momentum section is not too bad -- it slows down by some
33\%. This is in contrast to the Continuity and BCs sections which
slow down by factors of approximately three and two, respectively.

\begin{table}[!t]
\caption{The performance profile of the original and \psykal versions
  of nemolite2d (for 2000 time-steps of the $128^{2}$ domain and the
  Intel compiler).}
\label{TABLE_profile}
\centering
\begin{tabular}{l|c|c|c|c}
\hline
Section & \multicolumn{2}{c|}{Original} & \multicolumn{2}{c}{Vanilla \psykal} \\
        & Time (s) & \%  & Time (s) & \% \\
\hline
Continuity  & 0.10  & 2.7  & 0.33  & 6.6  \\
Momentum    & 3.00  & 84.5 & 4.05  & 79.5 \\
BCs         & 0.16  & 4.6  & 0.29  & 5.7  \\
Time-update & 0.29  & 8.1  & 0.41  & 8.1  \\
\hline
\end{tabular}
\end{table}

Beginning with the vanilla \psykal version, we then apply a series of
code transformations while obeying the \psykal separation of concerns,
{\it i.e.} optimisation is restricted to the middle, {PS}y layer and
leaves the kernel and algorithm layers unchanged. The aim of these
optimisations is to recover, as much as is possible, the performance
of the original version of the code. The transformations we have
performed and the reasons for them are described in the following sections.

\subsection{Addition of `safe-address' directives}

Much of the optimisations we have performed have been informed by the
diagnostic output produced by either the Cray or Intel compilers. Many
of the nemolite2d kernels contain conditional statements. These
statements are there to check whether {\it e.g.} the current grid
point is wet or neighbours a boundary point. A compiler is better able
to optimise such a loop if it can be sure that all array accesses
within the body of the loop are safe for every trip, irrespective of
the conditional statements. In its diagnostic output the Cray compiler
notes this with messages of the form:
\begin{verbatim}
ftn-6375 ftn: VECTOR File = nemolite2d.f90,
 Line = 448 
A loop starting at line 448 would benefit 
from "!dir$ safe_address".
\end{verbatim}
In order that we could safely add such directives we altered the field
module to allocate all field-data arrays with extents greater than
strictly required.  We were then able to safely add the safe\_address
before all of the loops where the Cray compiler indicated it might be
useful (the Momentum loops and some of the BC loops).

\subsection{In-line Momentum kernel bodies into middle-layer code}

From our previous work~\cite{shallow_psykal} on a similar code we know
that kernel in-lining is critical to obtaining performance with the
Gnu and Intel compilers. For the Gnu compiler, this is because it
cannot perform in-lining when routines are in separate source files
{\bf do we have evidence for this?}. Although the Intel compiler does
not have this limitation, we have found (both here and in our previous
work~\cite{shallow_psykal}) that the number of optimisations it
performs is reduced if it first has to in-line a routine.

The profiling data in Table~\ref{TABLE_profile} shows that it is the
Momentum section that accounts for the bulk of the model run-time.  We
therefore chose to in-line the two kernels (one for each of the $x$
and $y$ components of momentum) called within this section.

\subsection{Force SIMD vectorisation of the Momentum kernels using directives}

\subsection{Work around limitations related to derived types}

\subsection{In-line the Continuity kernel}

\subsection{In-line remaining kernels (BCs and time-update)}

\subsection{In-line field-copy operations}
